---
title: "Tidyverse ggplot tutorial"
author: "Monica Alexander"
date: "1/24/2020"
output: 
    html_document:
      number_sections: true
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Help

1. Don't be afraid to Google
2. R4DS: https://r4ds.had.co.nz/

Workflow:

1. Import
2. Tidy
3. Transforming, descriptive
4. Plot
5. Model
6. Repeat 3/4
7. Finish get PhD eat cake

# Tidy data

Tidy data:

1. each variable has its own column
2. each observation has its own row
3. each value has its own cell

```{r}
library(tidyverse)
```
This dataset is not in tidy form:

```{r}
d <- read_table("../data/CAN_Mx_1x1.txt", skip = 2)
d
```

How do we get there? `pivot_longer` and `pivot_wider`

```{r}
d <- pivot_longer(data = d, cols = Female:Total, names_to = "sex", values_to = "rate")
d
```

# Data transformation

First thing, the 'pipe': `%>%` Read it as "and then" (keyboard shortcut: Cmd+Shft+M)

Six functions that are useful to know:

- `select`
- `filter`
- `arrange`
- `mutate`
- `summarise` (`summarize`)
- `group_by`

The idea of the pipe is that you take your dataset **and then** you do stuff

```{r}
d <- read_rds("../data/infant.RDS")
d
```


## Select

Select is selecting a column. Get `cod`
```{r}
d %>% 
  select(cod) 
```
This is (almost) equivalent to the dollar sign:

```{r}
d$cod
```

You can get a vector from `select` by using `pull`

```{r}
d %>% 
  select(cod) %>% 
  pull()
```

You can also use `select` to get rid of columns:

```{r}
d %>% 
  select(-cod) 
```

## Filter

Filter allows you to select rows. Filter all males:

```{r}
d %>% 
  filter(sex=="M") 
```

Two conditions: male babies who died of SIDS (AND)

```{r}
d %>% 
  filter(sex=="M",cod=="sids")
```

Two conditions: male babies OR babies who died of SIDS

```{r}
d %>% 
  filter(sex=="M"|cod=="sids")
```

Build these up: using with `select`.  

```{r}
d %>% 
  filter(sex=="M", cod=="sids") %>% 
  select(aged)
```

`&` is the same as `,`

```{r}
d %>% 
  filter(cod=="sids"&sex=="M")
```

Filter a particular row (row 3):

```{r}
d %>% 
  filter(row_number()==3)
```

## Arrange

`arrange` is ordering data.

Arrange by `aged`

```{r}
d %>% 
  arrange(aged)
```

Arrange in descending order

```{r}
d %>% 
  arrange(-aged)
```

Arrange by multiple things (age of death and age of mom)

```{r}
d %>% 
  arrange(aged, mom_age)
```

What happens if we do two arranges:

```{r}
d %>% 
  arrange(aged) %>% 
  arrange(mom_age)
```

Same as

```{r}
d %>% 
  arrange(mom_age, aged)
```

## group_by

We might want to arrange by age of death **within** the sex group


```{r}
d %>% 
  group_by(sex) %>% 
  arrange(aged)
```

Say we want to arrange by aged of death, and then get the third row 

```{r}
d %>% 
  arrange(aged) %>% 
  filter(row_number()==3)
```

If we wanted to do this by sex of baby

```{r}
d %>% 
  group_by(sex) %>% 
  arrange(aged) %>% 
  filter(row_number()==3)
```

group by sex, arrange by age of death, get the first 3 rows:


```{r}
d %>% 
  group_by(sex) %>% 
  arrange(aged) %>% 
  filter(row_number() %in% 1:3)
```

Equivalently

```{r}
d %>% 
  group_by(sex) %>% 
  arrange(aged) %>% 
  slice(1:3)
```

## Mutate

`mutate` = make new variable

Firstly, let's change `gest` to days

```{r}
d <- d %>% 
  mutate(gest_days = gest*7)
d
```

Now, let's make a new variable that is total days in existence (aged+ gest_days)

```{r}
d %>% 
  mutate(tot_life = aged+gest_days)
```

## Summarise

`summarise` = new summary variables

Let's look at maximum of `mom_age`

```{r}
d %>% 
  summarise(max_mom_age = max(mom_age))
```

Check:

```{r}
d %>% 
  arrange(-mom_age)
```

Summarise is particularly powerful in conjunction with `group_by`.

Let's look at the maximum of mom age by `cod`


```{r}
d %>% 
  group_by(cod) %>% 
  summarise(max_mom_age = max(mom_age))
```

Let's look at mean of age at death by cod

```{r}
d %>% 
  group_by(cod) %>% 
  summarise(mean_age_of_death = round(mean(aged), 2)) %>% 
  arrange(-mean_age_of_death)
```

We can group by and summarise for multiple groups at a time:

mean age of death by cod and race of mother

```{r}
d %>% 
  group_by(cod, race) %>% 
  summarise(mean_age_of_death = round(mean(aged), 2)) %>% 
  arrange(-mean_age_of_death, race)
```

### Counting by group

Number of deaths by cod

```{r}
d %>% 
  group_by(cod) %>% 
  count()
```

### Calculate proportions

Often calculating proportions is a combination of summarise and mutate (and group by)

Let's calculate the proportion of deaths by cause. We're using `count` here, but it's sort of like `summarise`. 

Note here, that we needed to `ungroup()` the data before mutating. 

```{r}
d %>% 
  group_by(cod) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(prop = n/(sum(n)))
```

Highlight difference between mutate and summarise. 
Let's get the maximum age of death by cod group

```{r}
d %>% 
  group_by(cod) %>% 
  summarise(max_age_death = max(aged))
```

Could also do this as a mutate:

```{r}
d %>% 
  group_by(cod) %>% 
  mutate(max_age_death = max(aged)) %>% 
  arrange(cod)
```

Could be useful: use this as an intermediate step to get a new variable that is looking at age of death reelative to the maximum

```{r}
d %>% 
  group_by(cod) %>% 
  mutate(max_age_death = max(aged)) %>% 
  mutate(frac_max = aged/max_age_death)
```

# ggplot

ggplot is the plotting package that goes with tidyverse. 

In a similar to piping, it works in layers. Instead of piping ( %>% ) ggplot uses `+`. 

## Main features

1. data
2. aesthetics / mapping
3. type (`geom`)

Let's build up a histogram of age of death with increasing complexity. 

Starts with a grey box:

```{r}
ggplot(data = d, mapping = aes(x = aged))
```

We need to tell it what we want to plot. This is where `geom` comes in 

```{r}
ggplot(data = d, mapping = aes(x = aged)) +
  geom_histogram(fill = "blue")
```

Let's look at the histograms by race. This is an aesthetic. 

```{r}
ggplot(data = d, mapping = aes(x = aged, fill = race)) +
  geom_density(alpha = 0.5, color = NA)
```

Make it prettier?

```{r}
ggplot(data = d, mapping = aes(x = aged, fill = race)) +
  geom_density(alpha = 0.5, color = NA) + 
  labs(x = "Age of deaths (days)", 
       title = "Density of infant deaths", 
       subtitle = "because we can") +
  theme_bw()
  
```

Change the color

```{r}
ggplot(data = d, mapping = aes(x = aged, fill = cod)) +
  geom_density(alpha = 0.5, color = NA) + 
  labs(x = "Age of deaths (days)", 
       title = "Density of infant deaths", 
       subtitle = "because we can") +
  theme_bw() + 
  scale_fill_viridis_d()
```

## Facets

Facets == subplots

```{r}
ggplot(data = d, mapping = aes(x = aged, fill = cod)) +
  geom_density(alpha = 0.5, color = NA) + 
  labs(x = "Age of deaths (days)", 
       title = "Density of infant deaths", 
       subtitle = "because we can") +
  theme_bw() + 
  scale_fill_viridis_d() +
  facet_wrap(~race)
```

## Small things

1. You can pipe data into ggplot

```{r}
d %>% 
  group_by(cod, race) %>% 
  summarise(n = n()) %>% 
  group_by(race) %>% 
  mutate(prop = n/sum(n)) %>% 
  ggplot(aes(x = race, y = prop, fill = cod)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  scale_fill_viridis_d()
```

## Not ggplot but potentially useful

This graph above has a lot of categories. Let's change this into

- perinatal
- sid + maltreat
- everything else 

Use `case_when`

```{r}
d %>% 
  mutate(cod_aggregated = case_when(
    str_starts(cod, "peri") ~ "perintal",
    cod=="sids"|cod=="matreatment" ~ "sids+mal",
    TRUE ~ "other"
  )) %>% 
  group_by(cod_aggregated, race) %>% 
  summarise(n = n()) %>% 
  group_by(race) %>% 
  mutate(prop = n/sum(n)) %>% 
  ggplot(aes(x = race, y = prop, fill = cod_aggregated)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  scale_fill_viridis_d()
```

