---
title: "Help with Lab 2 Exercises"
author: "Monica Alexander"
date: "1/15/2020"
output: html_document
---
Very rough!

```{r}
library(tidyverse)
library(opendatatoronto)
library(janitor)
```

Get all the datasets available

```{r}
all_data <- list_packages(limit = 500)
all_data
```

Look at the campaigns 2014-2017 and download the data file

```{r}
list_package_resources("f6651a40-2f52-46fc-9e04-b760c16edd5c")
d <- get_resource("d99bb1f3-949a-4497-bb96-c93bbd203130")

```

`d` is a list of dataframes. find the majoral sheet and select that. Have a look at the dataframe

```{r}
names(d)
d <- d$`2_Mayor_Contributions_2014_election.xls`
d
```

Oh no! It look's like the column names are in the first row??? We don't want that. fix that up:

```{r}
colnames(d) <- d[1,]
d <- d[-1,]
?janitor::row_to_names # could also use this

d <- clean_names(d)
```

contributions need to be numeric

```{r}
d <- d %>% 
  mutate(amount = as.numeric(contribution_amount))
```


Visualize contributions

```{r}
ggplot(d, aes(x = amount)) + geom_histogram() + scale_x_log10()
```

look at top 10 amounts

```{r}
d %>% 
  arrange(-amount) %>% 
  select(contributors_name, relationship_to_candidate, candidate, amount)
```

```{r}
d %>% 
  filter(is.na(relationship_to_candidate)) %>% # remove those w/ relationship
  ggplot(aes(x = amount)) + geom_histogram() 
```

top candidates by total contributions

```{r}
d %>% 
  group_by(candidate) %>% 
  summarise(tot = sum(amount)) %>% 
  arrange(-tot)
```

