---
title: "GLM"
author: "Monica Alexander"
date: "January 22 2020"
output: 
    html_document:
      number_sections: true
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Today we are looking at data on infant deaths (deaths in the first year of life) in the US. The dataset `infant` contains information on all deaths to the 2012 birth cohort. For today, we are interested in investigating differences in neonatal deaths (i.e. deaths in the first month of life) and cause of death. 

## What to hand in

As with last week, please push your Rmd and compiled document (html or pdf) to GitHub. **The questions for this week are dispersed throughout the lab.**

## The dataset

Read it in and have a look to see what's in there. Variables are

- `sex`: sex of baby
- `aged`: age at death (in days)
- `race`: race of mother
- `gest`: gestation in weeks
- `ucod`: cause of death (ICD-10 code)
- `cod`: cause of death, descriptive groups
- `mom_age`: mother age in years
- `mom_age_group`: mother age group

```{r}
library(tidyverse)
library(here)
d <- read_rds(here("data", "infant.RDS"))
head(d)
```


# Descriptives

Let's create some new variables that will be useful:

- `neo_death`: equals 1 if the death occurred in the first 28 days
- `preterm`: equals 1 if gestational age is less than 37 weeks
- `cod_group`: reduced number of categories of cause of death

Also, removing the observations where we don't know gestational age or the mother's age.

```{r}
d <- d %>% 
  mutate(neo_death = ifelse(aged<=28, 1, 0),
         cod_group = case_when(
           str_starts(cod, "peri") ~ "perinatal",
           cod %in% c("other", "unknown") ~ "oth_unk",
           cod %in% c("sids", "maltreatment", "infection") ~ "exogenous",
           cod %in% c("resp", "heart") ~ "resp_heart",
           TRUE ~ cod
         ),
         preterm = ifelse(gest<37, 1, 0)) %>% 
  filter(gest<99, !is.na(mom_age_group))
```

## Distribution of gestational ages

Let's plot the distribution of gestational ages by race. It's quite bi-modal. Notice the difference in densities by race. 

```{r}
d %>% 
  ggplot(aes(gest, fill = race)) + geom_histogram(position = 'dodge', aes(y = ..density..))
```

## Question 1

Calculate the proportion of deaths that are neonatal by race and prematurity. Which group has the highest proportion of neonatal deaths?

```{r}
d %>% 
  group_by(race, preterm, neo_death) %>% 
  summarise(deaths = n()) %>% 
  group_by(race, preterm) %>% 
  mutate(prop = deaths/sum(deaths))
```


## Causes of death

Let's make `cod_group` a factor with congenital malformations as the reference.

```{r}
d <- d %>% 
  mutate(cod_group = factor(cod_group, levels = c("cong_mal", "perinatal", "resp_heart", "exogenous", "oth_unk"),
                         labels = c("cong_mal", "perinatal", "resp_heart", "exogenous", "oth_unk")))
```

The following code calculates the proportion of deaths by cause group, race, sex and prematurity

```{r}
prop_cause <- d %>% 
  group_by(race, preterm, sex, cod_group) %>% 
  summarise(n = n()) %>% 
  group_by(race, preterm, sex) %>% 
  mutate(prop = n/sum(n)) %>% 
  ungroup() %>% 
  mutate(preterm = ifelse(preterm==1,"pre-term", "full-term"))
```


## Question 2

Using the `prop_cause` above, filter to just look at female babies, and make a graph to help visualize differences in cause by race and prematurity. 


# Logistic regression

First, let's do logistic regression to explore differences in neonatal deaths. Here's a  model with prematurity, sex, race, and mom's age

```{r}
mod <- glm(neo_death~ preterm + sex + race + race:preterm + mom_age, data = d, family = binomial)
summary(mod)
```


## Question 3

Rerun the model above with instead of `mom_age`, include a new variable `mom_age_c` which centers mother's age around its mean. 

```{r}
mod <- glm(neo_death~ preterm + sex + race + race:preterm + mom_age_c, data = d, family = binomial)
summary(mod)
```

```{r}
preds <- predict(mod, pred_df, type = 'response')
pred_df %>% 
  cbind(preds) %>% 
  filter(sex=="F") %>% 
  ggplot(aes(factor(preterm), preds, col = race)) + geom_point()
```


## Question 4

Interpret the `preterm`, `race` and the interaction `preterm:race` coefficients. 

# Multinomial regression

Now let's do multinomial regression with cause of death as the outcome. We need to get the data in a different format to run the regression:

```{r}
d$mom_age_c <- d$mom_age - mean(d$mom_age)
d_wide <- d %>% 
  group_by(sex, race, cod_group, preterm,mom_age_c) %>% 
  summarise(deaths = n()) %>% 
  pivot_wider(names_from = cod_group, values_from = deaths) %>% 
  mutate_all(.funs = funs(ifelse(is.na(.), 0, .)))

d_wide$Y <- as.matrix(d_wide[,c("cong_mal","perinatal","resp_heart", "exogenous", "oth_unk")])
```

Now run the regression

```{r}
library(nnet)
mod2 <- multinom(Y ~ sex+race+ mom_age_c+ preterm, data = d_wide)
summary(mod2)
```

## Question 5

Using the `predict` function, find the predicted probabilities of each cause by race, sex and prematurity for the mothers of mean age. You can use this prediction dataframe to get all the combinations you need.

```{r}
pred_df <- tibble(preterm = c(rep(0, 4), rep(1, 4)),
       sex = rep(c(rep("F", 2), rep("M", 2)),2),
       race = rep(c("NHB", "NHW"), 4), 
       mom_age_c = 0)
```


```{r}
preds <- predict(mod2, newdata = pred_df, type = 'probs')

pred_df <- pred_df %>% cbind(preds)

```

## Question 6

Plot the predicted probabilities for female babies. 

## Question 7

What race/prematurity/ cause group has the highest probability? How does this compared to the observed proportion in the same group?